### 2019-08-09 ch2_dynbenchmark.tex
small fixes

\begin{figure}
	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/figure_2.pdf}
	Figure caption on next page $\rightarrow$
\end{figure}

\clearpage % insert a page break
\captionof{figure}{
	\textbf{A characterization of the 45 methods evaluated in this study and their overall evaluation results.}
	\textbf{a}, We characterized the methods according to the wrapper type, their required priors, whether the inferred topology is constrained by the algorithm (fixed) or a parameter (param), and the types of inferable topologies. The methods are grouped vertically based on the most complex trajectory type they can infer. \textbf{b}, The overall results of the evaluation on four criteria: accuracy using a reference trajectory on real and synthetic data, scalability with increasing number of cells and features, stability across dataset subsamples and quality of the implementation. Methods that errored on more than 50\% of the datasets are not included in this figure and are shown instead in Supplementary Fig. 2.
}
\label{fig:figure_2}

@@

\begin{figure}
	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/figure_3.pdf}
	Figure caption on next page $\rightarrow$
\end{figure}
\clearpage % insert a page break
\captionof{figure}{	
	\textbf{Detailed results of the four main evaluation criteria: accuracy, scalability, stability and usability.}
	\textbf{a,} The names of the methods, ordered as in Figure \ref{fig:figure_2}. \textbf{b}, Accuracy of trajectory inference methods across metrics, dataset sources and dataset trajectory types. The performance of a method is generally more stable across dataset sources, but very variable depending on the metric and trajectory type. \textbf{c}, Predicted execution times for varying numbers of cells and features (no. of cells $\times$ no. of features). Predictions were made by training a regression model after running each method on bootstrapped datasets with varying numbers of cells and features. k, thousands; m, millions; cor, correlation. \textbf{d}, Stability results by calculating the average pairwise similarity between models inferred across multiple runs of the same method. \textbf{e}, Usability scores of the tool and corresponding manuscript, grouped per category. Off-the-shelf methods were directly implemented in R and thus do not have a usability score. 
}
\label{fig:figure_3}

@@

\begin{figure}
	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/supfigure_3.pdf}
	Figure caption on next page $\rightarrow$
\end{figure}

\clearpage % insert a page break
\captionof{figure}{
	\textbf{Accuracy of trajectory inference methods.}
	\textbf{a} Overall score for all methods across 339 datasets, colored by the source of the datasets. Black line indicates the mean. \textbf{b} Similarity between the overall scores of all dataset sources, compared to real datasets with a gold standard, across all methods (n = 46, after filtering out methods that errored too frequently). Shown in the top left is the Pearson correlation. \textbf{c} Bias in the overall score towards trajectory types for all methods across 339 datasets. Black line indicates the mean. \textbf{d} Distributions of the difference in size between predicted and reference topologies. A positive difference means that the topology predicted by the method is more complex than the one in the reference.
}
\label{fig:supfigure_3}

@@

The high variability between datasets, together with the diversity in detected topologies between methods, could indicate some complementarity between the different methods. To test this, we calculated the likelihood of obtaining a top model when using only a subset of all methods. A top model in this case was defined as a model with an overall score of at least 95\% as the best model. On all datasets, using one method resulted in getting a top model about 27\% of the time. This increased up to 74\% with the addition of six other methods (Figure \ref{fig:figure_4}a). The result was a relatively diverse set of methods, containing both strictly linear or cyclic methods, and methods with a broad trajectory type range such as PAGA. We found similar indications of complementarity between the top methods on data containing only linear, bifurcation or multifurcating trajectories (Figure \ref{fig:figure_4}b), although in these cases less methods were necessary to obtain at least one top model for a given dataset. Altogether, this shows that there is considerable complementarity between the different methods and that users should try out a diverse set of methods on their data, especially when the topology is unclear a priori. Moreover, it also opens up the possibilities for new ensemble methods that utilize this complementarity.

\begin{figure}
	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/figure_4.pdf}
	\caption{
		\textbf{Complementarity between different trajectory inference methods.}
		\textbf{a}, We assessed the likelihood for different combinations of methods to lead to a ‘top model’ (defined as a model with an overall score of at least 95\% of the best model) when applied to all datasets. \textbf{b}, The likelihood for different combinations of methods to lead to a ‘top model’ was assessed separately on different trajectory types. For this figure, we did not include any methods requiring a cell grouping or a time course as prior information.
	}
	\label{fig:figure_4}
\end{figure}

@@

It is not only important that a method is able to infer an accurate model in a reasonable time frame, but also that it produces a similar model when given very similar input data. To test the stability of each method, we executed each method on ten different subsamples of the datasets (95\% of the cells, 95\% of the features), and calculated the average similarity between each pair of models using the same scores used to assess the accuracy of a trajectory (Figure \ref{fig:figure_3}d).

@@

	\begin{tabular}{p{2.5cm}p{7.5cm}p{3cm}}
		\toprule
		Aspect & Items & References \\ \midrule

@@

We found that the performance of a method can be very variable between datasets, and therefore included a large set of both real and synthetic data within our evaluation, leading to a robust overall ranking of the different methods. However, ‘good-yet-not-the-best’ methods \cite{norel_selfassessmenttrap_2011} can still provide a very valuable contribution to the field, especially if they make use of novel algorithms, return a more scalable solution or provide a unique insight in specific use cases. This is also supported by our analysis of method complementarity. Some examples for the latter include PhenoPath, which can include additional covariates in its model, ouija, which returns a measure of uncertainty of each cell’s position within the trajectory, and StemID, which can infer the directionality of edges within the trajectory.

@@

\begin{itemize}
	\item \textbf{CALISTA}: We assigned the cells to the branch at which the sum of the cluster probabilities of two connected milestones was the highest. The cluster probabilities of the two selected milestones were then used as milestone percentages. This was then processed as a type 1, direct, method.
	\item \textbf{DPT}: We projected the cells onto the cluster network, consisting of a central milestone (this cluster contained the cells that were assigned to the ‘unknown’ branch) and three terminal milestones, each corresponding to a tip point. This was then processed as a type 1, direct, method.
	\item \textbf{Sincell}: To constrain the number of milestones this method creates, we merged two cell clusters iteratively until the percentage of leaf nodes was below a certain cutoff, with the default cutoff set to 25\%. This was then processed as a type 7, cell graph, method.
	\item \textbf{SLICE}: As discussed in the vignette of SLICE (\href{https://research.cchmc.org/pbge/slice.html}{https://research.cchmc.org/pbge/slice.html}), we ran principal curves one by one for every edge detected by SLICE. This was then processed as a type 1, direct, method.
	\item \textbf{MFA}: We used the branch assignment as state probabilities, which together with the global pseudotime were processed as a type 4, end state probabilities, method.
	\item \textbf{URD}: We extracted the pseudotime of a cell within each branch using the y positions in the tree layout. This was then further processed as a type 1, direct, method.

@@

\begin{itemize}
	\item \textbf{dyngen}: simulations of gene regulatory networks, available at \href{https://github.com/dynverse/dyngen}{https://github.com/dynverse/dyngen}
	\item \textbf{dyntoy}: random gradients of expression in the reduced space, available at \href{https://github.com/dynverse/dyntoy}{https://github.com/dynverse/dyntoy}
	\item \textbf{PROSSTT}: expression is sampled from a linear model that depends on pseudotime \cite{papadopoulosPROSSTTProbabilisticSimulation2018}
	\item \textbf{Splatter}: simulations of non-linear paths between different expression states \cite{zappia_splattersimulationsinglecell_2017}

For every simulator, we took great care to make the datasets as realistic as possible. To do this, we extracted several parameters from all real datasets. We calculated the number of differentially expressed features within a trajectory using a two-way Mann–Whitney U test between every pair of cell groups. These values were corrected for multiple testing using the Benjamini-Hochberg procedure (FDR < 0.05) and we required that a gene was expressed in at least 5\% of cells, and had at least a fold-change of 2. We also calculated several other parameters, such as drop-out rates and library sizes using the Splatter package \cite{zappia_splattersimulationsinglecell_2017}. These parameters were then given to the simulators when applicable, as described for each simulator below. Not every real dataset was selected to serve as a reference for a synthetic dataset. Instead, we chose a set of ten distinct reference real datasets by clustering all the parameters of each real dataset, and used the reference real datasets at the cluster centers from a pam clustering (with $k = 10$, implemented in the R cluster package) to generate synthetic data.

@@

From these module networks we generated gene regulatory networks in two steps: the main regulatory network was first generated, and extra target genes from real regulatory networks  were added. For each dataset, we used the same number of genes as were differentially expressed in the real datasets. 5\% of the genes were assigned to be part of the main regulatory network, and were randomly distributed among all modules (with at least one gene per module). We sampled edges between these individual genes (according to the module network) using a uniform distribution between 1 and the number of possible targets in each module. To add additional target genes to the network, we assigned every regulator from the network to a real regulator in a real network (from regulatory circuits \cite{marbach_tissuespecificregulatorycircuits_2016}), and extracted for every regulator a local network around it using personalized pagerank (with damping factor set to 0.1), as implemented in the page\_rank function of the \textit{igraph} package.

@@

To simulate the gene regulatory network, we used a system of differential equations similar to those used in evaluations of gene regulatory network inference methods \cite{marbach_wisdom_2012}. In this model, the changes in gene expression ($x_i$) and protein expression ($y_i$) are modeled using ordinary differential equations \cite{schaffter_genenetweaversilicobenchmark_2011} (ODEs):

@@

Similar to GeneNetWeaver \cite{schaffter_genenetweaversilicobenchmark_2011}, we sample the different parameters from random distributions, defined as follows. $e$ defines whether a transcription factor activates (1) or represses (-1), as defined within the regulatory network network.

@@

We used a standard single-cell RNA-seq preprocessing pipeline that applies parts of the scran and scater Bioconductor packages \cite{lun_stepbystepworkflowlowlevel_2016}. The advantages of this pipeline are that it works both with and without spike-ins, and it includes a harsh cell filtering that looks at abnormalities in library sizes, mitochondrial gene expression and the number of genes expressed using median absolute deviations (which we set to 3). We required that a gene was expressed in at least 5\% of the cells and that it should have an average expression higher than 0.02. Furthermore, we used the pipeline to select the most highly variable genes, using a false discovery rate of 5\% and a biological component higher than 0.5. As a final filter, we removed both all-zero genes and cells until convergence.

@@

The wcorfeatures assesses whether the same differentially expressed features are found using the predicted trajectory as in the known trajectory. To calculate this metric, we used Random Forest regression (implemented in the R ranger package \cite{wright_rangerfastimplementation_2017}), to predict expression values of each gene, based on the geodesic distances of a cell to each milestone. We then extracted feature importance values for each feature and calculated the similarity of the feature importances using a weighted Pearson correlation, weighted by the feature importance in the reference dataset to give more weight to large differences. As hyperparameters we set the number of trees to 10,000 and the number of features on which to split to 1\% of all available features. We illustrate this metric and assess the effect of its hyperparameters in Supplementary Note 1.

@@

To assess the complementarity between different methods, we first calculated for every method and dataset whether the overall score was equal to or higher than 95\% of the best overall score for that particular dataset. We then calculated for every method the weighted percentage of datasets that fulfilled this rule, weighted similarly as in the benchmark aggregation, and chose the best method. We iteratively added new methods until all methods were selected. For this analysis, we did not include any methods that require any strong prior information and only included methods that could detect the trajectory types present in at least one of the datasets.

@@

In the ideal case, a method should produce a similar trajectory, even when the input data is slightly different. However, running the method multiple times on the same input data would not be the ideal approach to assess its stability, given that a lot of tools are artificially deterministic by internally resetting the pseudorandom number generator (for example, using the `set.seed` function in R or the `random.seed` function in numpy). To assess the stability of each method, we therefore selected a number of datasets, which consisted of 25\% of the datasets accounting for 15\% of the total runtime, chosen such that after aggregation the overall scores still has > 0.99 correlation with the original overall ranking. We subsampled each dataset 10 times with 95\% of the original cells and 95\% of the original features. We ran every method on each of the bootstraps, and assessed the stability by calculating the benchmarking scores between each pair of subsequent models (run $i$ is compared to run $i+1$). For the cordist and F1branches, we only used the intersection between the cells of two datasets, while the intersection of the features was used for the wcorfeatures.

@@

\begin{figure}
	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/snote1fig_5.pdf}
	\caption{
		\textbf{Determination of cell waypoints} \textbf{a)} Illustration of the stratified cell sampling using an example dataset (top). Each milestone, edge between two milestones and region of delayed commitment is seen as a collection of cells (middle), and the number of waypoints (100 in this case) are divided over each of these collection of cells (bottom). \textbf{b)} Accuracy versus time to calculate $\textit{cor}_{\textit{dist}}$. Shown are distributions over 100 random waypoint samples. The upper whisker of the boxplot extends from the hinge (75\% percentile) to the largest value, no further than 1.5$\times$ the IQR of the hinge. The lower whisker extends from the hinge (25\% percentile) to the smallest value, at most 1.5$\times$ the IQR of the hinge.
	}
	\label{fig:snote1fig_5}
\end{figure}

@@

Random forest regression has two main hyperparameters. The number of trees to be fitted (num\_tree parameter) was fixed to 10000 to provide accurate and stable estimates of the feature importance (Figure \ref{fig:snote1fig_9}. The number of features on which can be split (mtry parameter) was set to 1\% of all available features (instead of the default square-root of the number of features), as to make sure that predictive but highly correlated features, omnipresent in transcriptomics data, are not suppressed in the ranking.

@@

For most datasets, only a limited number of features will be differentially expressed in the trajectory. For example, in the dataset used in Figure \ref{fig:snote1fig_9} only the top 10\%-20\% show a clear pattern of differential expression. The correlation will weight each of these features equally, and will therefore give more weight to the bottom, irrelevant features. To prioritise the top differentially expressed features, we also implemented the $\textit{wcor}_{\textit{features}}$, which will weight the correlation using the feature importance scores in the reference so that the top features have relatively more impact on the score (Figure \ref{fig:snote1fig_10}).

>>>>

\begin{figure}
	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/figure_2.pdf}
	\caption{
		\textbf{A characterization of the 45 methods evaluated in this study and their overall evaluation results.}
		\textbf{a}, We characterized the methods according to the wrapper type, their required priors, whether the inferred topology is constrained by the algorithm (fixed) or a parameter (param), and the types of inferable topologies. The methods are grouped vertically based on the most complex trajectory type they can infer. \textbf{b}, The overall results of the evaluation on four criteria: accuracy using a reference trajectory on real and synthetic data, scalability with increasing number of cells and features, stability across dataset subsamples and quality of the implementation. Methods that errored on more than 50$\%$ of the datasets are not included in this figure and are shown instead in Supplementary Fig. 2.
	}
	\label{fig:figure_2}
\end{figure}

@@

\begin{figure}
	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/figure_3.pdf}
	\caption{	
		\textbf{Detailed results of the four main evaluation criteria: accuracy, scalability, stability and usability.}
		\textbf{a,} The names of the methods, ordered as in Figure \ref{fig:figure_2}. \textbf{b}, Accuracy of trajectory inference methods across metrics, dataset sources and dataset trajectory types. The performance of a method is generally more stable across dataset sources, but very variable depending on the metric and trajectory type. \textbf{c}, Predicted execution times for varying numbers of cells and features (no. of cells $\times$ no. of features). Predictions were made by training a regression model after running each method on bootstrapped datasets with varying numbers of cells and features. k, thousands; m, millions; cor, correlation. \textbf{d}, Stability results by calculating the average pairwise similarity between models inferred across multiple runs of the same method. \textbf{e}, Usability scores of the tool and corresponding manuscript, grouped per category. Off-the-shelf methods were directly implemented in R and thus do not have a usability score. 
	}
	\label{fig:figure_3}
\end{figure}

@@

\begin{figure}
	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/supfigure_3.pdf}
	\caption{
		\textbf{Accuracy of trajectory inference methods.}
		\textbf{a} Overall score for all methods across 339 datasets, colored by the source of the datasets. Black line indicates the mean. \textbf{b} Similarity between the overall scores of all dataset sources, compared to real datasets with a gold standard, across all methods (n = 46, after filtering out methods that errored too frequently). Shown in the top left is the Pearson correlation. \textbf{c} Bias in the overall score towards trajectory types for all methods across 339 datasets. Black line indicates the mean. \textbf{d} Distributions of the difference in size between predicted and reference topologies. A positive difference means that the topology predicted by the method is more complex than the one in the reference.
	}
	\label{fig:supfigure_3}
\end{figure}

@@

The high variability between datasets, together with the diversity in detected topologies between methods, could indicate some complementarity between the different methods. To test this, we calculated the likelihood of obtaining a top model when using only a subset of all methods. A top model in this case was defined as a model with an overall score of at least 95$\%$ as the best model. On all datasets, using one method resulted in getting a top model about 27$\%$ of the time. This increased up to 74$\%$ with the addition of six other methods (Figure \ref{fig:figure_4}a). The result was a relatively diverse set of methods, containing both strictly linear or cyclic methods, and methods with a broad trajectory type range such as PAGA. We found similar indications of complementarity between the top methods on data containing only linear, bifurcation or multifurcating trajectories (Figure \ref{fig:figure_4}b), although in these cases less methods were necessary to obtain at least one top model for a given dataset. Altogether, this shows that there is considerable complementarity between the different methods and that users should try out a diverse set of methods on their data, especially when the topology is unclear a priori. Moreover, it also opens up the possibilities for new ensemble methods that utilize this complementarity.

\begin{figure}
	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/figure_4.pdf}
	\caption{
		\textbf{Complementarity between different trajectory inference methods.}
		\textbf{a}, We assessed the likelihood for different combinations of methods to lead to a ‘top model’ (defined as a model with an overall score of at least 95$\%$ of the best model) when applied to all datasets. \textbf{b}, The likelihood for different combinations of methods to lead to a ‘top model’ was assessed separately on different trajectory types. For this figure, we did not include any methods requiring a cell grouping or a time course as prior information.
	}
	\label{fig:figure_4}
\end{figure}

@@

It is not only important that a method is able to infer an accurate model in a reasonable time frame, but also that it produces a similar model when given very similar input data. To test the stability of each method, we executed each method on ten different subsamples of the datasets (95$\%$ of the cells, 95$\%$ of the features), and calculated the average similarity between each pair of models using the same scores used to assess the accuracy of a trajectory (Figure \ref{fig:figure_3}d).

@@

	\begin{tabular}{p{3.5cm}p{9cm}p{2.5cm}}
		\toprule
		Aspect & Items & References \\ \midrule

@@

We found that the performance of a method can be very variable between datasets, and therefore included a large set of both real and synthetic data within our evaluation, leading to a robust overall ranking of the different methods. However, ‘good-yet-not-the-best’ methods \cite{Norel2011} can still provide a very valuable contribution to the field, especially if they make use of novel algorithms, return a more scalable solution or provide a unique insight in specific use cases. This is also supported by our analysis of method complementarity. Some examples for the latter include PhenoPath, which can include additional covariates in its model, ouija, which returns a measure of uncertainty of each cell’s position within the trajectory, and StemID, which can infer the directionality of edges within the trajectory.

@@

\begin{itemize}
	\item \textbf{CALISTA}: We assigned the cells to the branch at which the sum of the cluster probabilities of two connected milestones was the highest. The cluster probabilities of the two selected milestones were then used as milestone percentages. This was then processed as a type 1, direct, method.
	\item \textbf{DPT}: We projected the cells onto the cluster network, consisting of a central milestone (this cluster contained the cells that were assigned to the ‘unknown’ branch) and three terminal milestones, each corresponding to a tip point. This was then processed as a type 1, direct, method.
	\item \textbf{Sincell}: To constrain the number of milestones this method creates, we merged two cell clusters iteratively until the percentage of leaf nodes was below a certain cutoff, with the default cutoff set to 25$\%$. This was then processed as a type 7, cell graph, method.
	\item \textbf{SLICE}: As discussed in the vignette of SLICE (\href{https://research.cchmc.org/pbge/slice.html}{https://research.cchmc.org/pbge/slice.html}), we ran principal curves one by one for every edge detected by SLICE. This was then processed as a type 1, direct, method.
	\item \textbf{MFA}: We used the branch assignment as state probabilities, which together with the global pseudotime were processed as a type 4, end state probabilities, method.
	\item \textbf{URD}: We extracted the pseudotime of a cell within each branch using the y positions in the tree layout. This was then further processed as a type 1, direct, method.

@@

\begin{itemize}
	\item \textbf{dyngen}: simulations of gene regulatory networks, available at \href{https://github.com/dynverse/dyngen}{https://github.com/dynverse/dyngen}
	\item \textbf{dyntoy}: random gradients of expression in the reduced space, available at \href{https://github.com/dynverse/dyntoy}{https://github.com/dynverse/dyntoy}
	\item \textbf{PROSSTT}: expression is sampled from a linear model that depends on pseudotime \cite{papadopoulosPROSSTTProbabilisticSimulation2018}
	\item \textbf{Splatter}: simulations of non-linear paths between different expression states \cite{zappia_splattersimulationsinglecell_2017}
\end{itemize}

For every simulator, we took great care to make the datasets as realistic as possible. To do this, we extracted several parameters from all real datasets. We calculated the number of differentially expressed features within a trajectory using a two-way Mann–Whitney U test between every pair of cell groups. These values were corrected for multiple testing using the Benjamini-Hochberg procedure (FDR < 0.05) and we required that a gene was expressed in at least 5$\%$ of cells, and had at least a fold-change of 2. We also calculated several other parameters, such as drop-out rates and library sizes using the Splatter package \cite{zappia_splattersimulationsinglecell_2017}. These parameters were then given to the simulators when applicable, as described for each simulator below. Not every real dataset was selected to serve as a reference for a synthetic dataset. Instead, we chose a set of ten distinct reference real datasets by clustering all the parameters of each real dataset, and used the reference real datasets at the cluster centers from a pam clustering (with $k = 10$, implemented in the R cluster package) to generate synthetic data.

@@

From these module networks we generated gene regulatory networks in two steps: the main regulatory network was first generated, and extra target genes from real regulatory networks  were added. For each dataset, we used the same number of genes as were differentially expressed in the real datasets. 5$\%$ of the genes were assigned to be part of the main regulatory network, and were randomly distributed among all modules (with at least one gene per module). We sampled edges between these individual genes (according to the module network) using a uniform distribution between 1 and the number of possible targets in each module. To add additional target genes to the network, we assigned every regulator from the network to a real regulator in a real network (from regulatory circuits \cite{marbach_tissuespecificregulatorycircuits_2016}), and extracted for every regulator a local network around it using personalized pagerank (with damping factor set to 0.1), as implemented in the page\_rank function of the \textit{igraph} package.

@@

To simulate the gene regulatory network, we used a system of differential equations similar to those used in evaluations of gene regulatory network inference methods \cite{marbach_wisdom_2012}. In this model, the changes in gene expression ($x_i$) and protein expression ($y_i$) are modeled using ordinary differential equations \cite{schaffter2011} (ODEs):

@@

Similar to GeneNetWeaver \cite{schaffter2011}, we sample the different parameters from random distributions, defined as follows. $e$ defines whether a transcription factor activates (1) or represses (-1), as defined within the regulatory network network.

@@

We used a standard single-cell RNA-seq preprocessing pipeline that applies parts of the scran and scater Bioconductor packages \cite{lun_stepbystepworkflowlowlevel_2016}. The advantages of this pipeline are that it works both with and without spike-ins, and it includes a harsh cell filtering that looks at abnormalities in library sizes, mitochondrial gene expression and the number of genes expressed using median absolute deviations (which we set to 3). We required that a gene was expressed in at least 5$\%$ of the cells and that it should have an average expression higher than 0.02. Furthermore, we used the pipeline to select the most highly variable genes, using a false discovery rate of 5$\%$ and a biological component higher than 0.5. As a final filter, we removed both all-zero genes and cells until convergence.

@@

The wcorfeatures assesses whether the same differentially expressed features are found using the predicted trajectory as in the known trajectory. To calculate this metric, we used Random Forest regression (implemented in the R ranger package \cite{wright_rangerfastimplementation_2017}), to predict expression values of each gene, based on the geodesic distances of a cell to each milestone. We then extracted feature importance values for each feature and calculated the similarity of the feature importances using a weighted Pearson correlation, weighted by the feature importance in the reference dataset to give more weight to large differences. As hyperparameters we set the number of trees to 10,000 and the number of features on which to split to 1$\%$ of all available features. We illustrate this metric and assess the effect of its hyperparameters in Supplementary Note 1.

@@

To assess the complementarity between different methods, we first calculated for every method and dataset whether the overall score was equal to or higher than 95$\%$ of the best overall score for that particular dataset. We then calculated for every method the weighted percentage of datasets that fulfilled this rule, weighted similarly as in the benchmark aggregation, and chose the best method. We iteratively added new methods until all methods were selected. For this analysis, we did not include any methods that require any strong prior information and only included methods that could detect the trajectory types present in at least one of the datasets.

@@

In the ideal case, a method should produce a similar trajectory, even when the input data is slightly different. However, running the method multiple times on the same input data would not be the ideal approach to assess its stability, given that a lot of tools are artificially deterministic by internally resetting the pseudorandom number generator (for example, using the `set.seed` function in R or the `random.seed` function in numpy). To assess the stability of each method, we therefore selected a number of datasets, which consisted of 25$\%$ of the datasets accounting for 15$\%$ of the total runtime, chosen such that after aggregation the overall scores still has > 0.99 correlation with the original overall ranking. We subsampled each dataset 10 times with 95$\%$ of the original cells and 95$\%$ of the original features. We ran every method on each of the bootstraps, and assessed the stability by calculating the benchmarking scores between each pair of subsequent models (run $i$ is compared to run $i+1$). For the cordist and F1branches, we only used the intersection between the cells of two datasets, while the intersection of the features was used for the wcorfeatures.

@@

\begin{figure}
	\centering\includegraphics[width=\linewidth]{fig/dynbenchmark/snote1fig_5.pdf}
	\caption{
		\textbf{Determination of cell waypoints} \textbf{a)} Illustration of the stratified cell sampling using an example dataset (top). Each milestone, edge between two milestones and region of delayed commitment is seen as a collection of cells (middle), and the number of waypoints (100 in this case) are divided over each of these collection of cells (bottom). \textbf{b)} Accuracy versus time to calculate $\textit{cor}_{\textit{dist}}$. Shown are distributions over 100 random waypoint samples. The upper whisker of the boxplot extends from the hinge (75$\%$ percentile) to the largest value, no further than 1.5$\times$ the IQR of the hinge. The lower whisker extends from the hinge (25$\%$ percentile) to the smallest value, at most 1.5$\times$ the IQR of the hinge.
	}
	\label{fig:snote1fig_5}
\end{figure}

@@

Random forest regression has two main hyperparameters. The number of trees to be fitted (num\_tree parameter) was fixed to 10000 to provide accurate and stable estimates of the feature importance (Figure \ref{fig:snote1fig_9}. The number of features on which can be split (mtry parameter) was set to 1$\%$ of all available features (instead of the default square-root of the number of features), as to make sure that predictive but highly correlated features, omnipresent in transcriptomics data, are not suppressed in the ranking.

@@

For most datasets, only a limited number of features will be differentially expressed in the trajectory. For example, in the dataset used in Figure \ref{fig:snote1fig_9} only the top 10$\%$-20$\%$ show a clear pattern of differential expression. The correlation will weight each of these features equally, and will therefore give more weight to the bottom, irrelevant features. To prioritise the top differentially expressed features, we also implemented the $\textit{wcor}_{\textit{features}}$, which will weight the correlation using the feature importance scores in the reference so that the top features have relatively more impact on the score (Figure \ref{fig:snote1fig_10}).
